Algebraic Principles for Rely-Guarantee Style Concurrency Verification Tools

### 2013-11-12 Paper/paper.tex
Almost finished paper 2

\newtheorem{lem}{Lemma}

@@

For logics of programs, notions of iteration or recursion are
essential. A \emph{Kleene algebra} is a dioid expaned with a star
operation which satisfies the \emph{left unfold axiom} $1 + x\cdot
x^*\le x^*$ and the \emph{left} and \emph{right induction axiom} $z+x
\cdot y \le y\Rightarrow x^*\cdot z \le y$ and $z + y\cdot x \le y
z\cdot \Rightarrow x^* \le y$. It follows that $1+x\cdot x^* = x^*$
and that the right unfold axiom $1+x^*\cdot x=x^*$ is derivable as
well. Thus iteration $x^\ast$ is modelled as the least fixpoint of the
function $\lambda y. 1+x\cdot y$, which is the same as the least
fixpoint of $\lambda y.1+ y\cdot x$. A \emph{commutative Kleene
  algebra} is a Kleene algebra in which multiplication is commutative.

@@

A rely-guarantee algebra is a structure
$(K,C,+,\sqcap,\cdot,\|,^\star,0,1)$, where $(K,+,\sqcap)$ is a
distributive lattice, $(K,+,\cdot,\|,0,1)$ is a trioid and
$(K,+,\cdot,^\star,0,1)$ is a Kleene algebra. Note that we do not
consider the parallel star. $C$ is a distinguished subset of rely and
guarantee conditions or \emph{interference constraints} which satisfy
the following axioms
\begin{align}
r\|r &\le r, \label{rg1}\\
r &\le r\|r', \label{rg2}\\
r\|xy &= (r\|x)(r\|y), \label{rg3}\\
r\|x^+ &\le (r\|x)^+ \label{rg4}.
\end{align}
By convention, we use $r$ and $g$ to refer to elements of $C$,
depending on whether they are used as relies or guarantees, and
$x,y,z$ for arbitrary elements of $K$. The operators $\|$ and $\sqcap$
must be closed with respect to $C$.

Axiom (\ref{rg1}) states that interference from a rely condition being
run twice in parallel is no different from just the interference from
that rely condition begin run once in parallel. Axiom (\ref{rg2})
states that interference from a single rely condition is less than
interference from itself and another rely condition. Axiom (\ref{rg3})
allows a rely or guarantee to be split across sequential
programs. Axiom (\ref{rg4}) is similar to Axiom (\ref{rg3}) in intent,
except it deals with finite iteration.

@@

Axioms (\ref{rg1}), (\ref{rg2}) and (\ref{rg3}) are independent. To
show this, we use Isabelle's
\emph{Nitpick}~\cite{blanchette_nitpick:_2010} counterexample
generator to construct models which violate each particular axiom
while satisfying all others.

Axiom (\ref{rg4}) can be derived from (\ref{rg3}) by a simple
inductive argument in quantales where the plus operation can be
defined as a sum of powers $x^+=\sum_{i\ge 1} x^i$ where $x^1=x$
and $x^{i+1}=x\cdot x^i$. In first-order Kleene algebras, however, it
is independent, but it is impossible to find a counterexample with Nitpick because all such counterexamples are finite, but all finite Kleene algebras are a forteriori quantales.

Jones quintuples can be encodeded in this setting as
\begin{align}
r, g \vdash \{p\} c \{q\} \iff p(r\|c) \le q \land c \le g. \label{quin}
\end{align}
With this encoding we can derive the standard rely-guarantee inference
rules, as shown in Figure \ref{fig:rgrules}. Thus equations
(\ref{rg1}) to (\ref{rg4}), which are all necessary to derive these
rules, somehow represent a minimal set of axioms from which these
inference rules can be derived.

By using these residuals quintuples can be encoded in the following way,
which is equivalent to the encoding in Equation (\ref{quin}).
\begin{align}
r, g \vdash \{p\} c \{q\} \iff c \le r/(p \rightarrow q) \sqcap g \label{refine}.
\end{align}
This encoding allows us to think in terms of program refinement~\cite{hayes_refining_2013}, as
$r/(p \rightarrow q) \sqcap g$ defines the weakest program that when
placed in parallel with interference from $r$, and guaranteeing
interference at most $g$, goes from $p$ to $q$---a generic
specification for a concurrent program.

\begin{figure}[tbh]
\centering
\begin{prooftree}
\RightLabel{Skip}
\AxiomC{$pr \le p$}
\UnaryInfC{$r, g \vdash \{p\}1\{p\}$}
\end{prooftree}

@@

@@

While the algebra in section \ref{sec:RG} is adequate for deriving the
standard inference rules, it's equality is too strong to capture many
interesting statements about concurrent programs. Consider the
congruence rule for parallel composition, which is inherent in the
algebraic approach:
\begin{align*}
x = y \implies x\|z = y\|z.
\end{align*}
This can be read as follows; if $x$ and $y$ are equal, then they must
be equal under all possible interference from any arbitrary $z$. At
first sight, this might seem to preclude any reasoning about
interference in a fine-grained way using purely algebra. This is not
the case, but breaking such inherent compositionality in just the
right way to capture interesting properties of concurrency requires
extra work.

The most obvious way we can acheive this is by expanding our
rely-guarantee algebra with an additional function $\pi$ and redefining
our quintuples as,
\begin{align*}
r, g \vdash \{p\} c \{q\} \iff \pi(p(r\|c)) \le \pi(q) \land c \le g.
\end{align*}
Since for any operator $\bullet$, it is not required that
\begin{align*}
\pi(x) = \pi(y) \implies \pi(x \bullet z) = \pi(y \bullet z),
\end{align*}
we can `break' compositionality in just the right way, provided we
choose the correct properties for $\pi$. For example, if we consider
the following seven axioms
\begin{align}
\pi(\pi(x)) &= \pi(x), \label{pi1}\\
x \le y \implies \pi(x) &\le \pi(y), \label{pi2}\\
\pi(x) &\le x, \label{pi3}\\
\pi(x^*) &\le \pi(\pi(x)^*), \label{pi4}\\
\pi(xy) &\le \pi(\pi(x)\pi(y)), \label{pi5}\\
\pi(x \sqcap y) &= \pi(x) \sqcap \pi(y). \label{pi6}
\end{align}

Axioms (\ref{pi1}) to (\ref{pi3}) show that $\pi$ is an interior
operator, while (\ref{pi4}) and (\ref{pi5}) allow us to introduce
$\pi$ as needed around sequential composition and finite iteration,
which allows us to substitute under those operators as required---we
only want to break compositionality for $\|$.

The above axioms indicate that we could define $\pi$ as a function
$\lambda x.\; x \sqcap c$ where $c$ satisfies
\begin{align}
x^* \sqcap c &\le (x \sqcap c)^* \sqcap c, \label{con1}\\
xy \sqcap c &\le (x \sqcap c)(y \sqcap c) \sqcap c. \label{con2}
\end{align}

As such we redefine our rely-guarantee algebra from section
\ref{sec:RG} as a structure $(K,RG,+,\sqcap,\cdot,\|,^\star,0,1,c)$
which, in addition to the rules in section \ref{sec:RG} satisfies both
(\ref{con1}) and (\ref{con2}).

All the rules in Figure \ref{fig:rgrules} can be derived with this
algebra, and moreover their proofs remain the same, mutatis
mutandis.

@@

CNPq EPSRC

>>>>

\newtheorem{lem}{Lemma}
\newtheorem{thm}{Theorem}

@@

For logics of programs, notions of iteration or recursion are
essential. A \emph{Kleene algebra} is a dioid expaned with a star
operation which satisfies the \emph{left unfold axiom} $1 + x\cdot
x^*\le x^*$ and the \emph{left} and \emph{right induction axiom} $z+x
\cdot y \le y\Rightarrow x^*\cdot z \le y$ and $z + y\cdot x \le y
\Rightarrow z\cdot x^* \le y$. It follows that $1+x\cdot x^* = x^*$
and that the right unfold axiom $1+x^*\cdot x=x^*$ is derivable as
well. Thus iteration $x^\ast$ is modelled as the least fixpoint of the
function $\lambda y. 1+x\cdot y$, which is the same as the least
fixpoint of $\lambda y.1+ y\cdot x$. A \emph{commutative Kleene
  algebra} is a Kleene algebra in which multiplication is commutative.

@@

A rely-guarantee algebra is a structure
$(K,I,+,\sqcap,\cdot,\|,^\star,0,1)$, where $(K,+,\sqcap)$ is a
distributive lattice, $(K,+,\cdot,\|,0,1)$ is a trioid and
$(K,+,\cdot,^\star,0,1)$ is a Kleene algebra. Note that we do not
consider the parallel star. $I$ is a distinguished subset of rely and
guarantee conditions or \emph{interference constraints} which satisfy
the following axioms
\begin{align}
r\|r &\le r, \label{rg1}\\
r &\le r\|r', \label{rg2}\\
r\|xy &= (r\|x)(r\|y), \label{rg3}\\
r\|x^+ &\le (r\|x)^+ \label{rg4}.
\end{align}
By convention, we use $r$ and $g$ to refer to elements of $I$,
depending on whether they are used as relies or guarantees, and
$x,y,z$ for arbitrary elements of $K$. The operators $\|$ and $\sqcap$
must be closed with respect to $I$.

Axiom (\ref{rg1}) states that interference from a constraint being run
twice in parallel is no different from just the interference from that
constraint begin run once in parallel. Axiom (\ref{rg2}) states that
interference from a single constraint is less than interference from
itself and another interference constraint. Axiom (\ref{rg3}) allows
an interference constraint to be split across sequential
programs. Axiom (\ref{rg4}) is similar to Axiom (\ref{rg3}) in intent,
except it deals with finite iteration.

@@

\begin{thm}
  Axioms (\ref{rg1}), (\ref{rg2}) and (\ref{rg3}) are independent.
\end{thm}
\begin{proof}
  To show this, we use Isabelle's
  \emph{Nitpick}~\cite{blanchette_nitpick:_2010} counterexample
  generator to construct models which violate each particular axiom
  while satisfying all others.
\end{proof}

\begin{thm}
Axiom (\ref{rg4}) can be derived from (\ref{rg3}) by a simple
inductive argument in a quantale where $\|$ distributes over arbitrary suprema.
\end{thm}
\begin{proof}
  In a quantale $x^+$ can be defined as a sum of powers
  $x^+=\sum_{i\ge 1} x^i$ where $x^1=x$ and $x^{i+1}=x\cdot x^i$. By
  induction on w$i$ we get $r\|x^i = (r\|x)^i$, hence
  \begin{align*}
    r\|x^+ = r\|\sum_{i\ge 1} x^i = \sum_{i\ge 1} r\|x^i = \sum_{i\ge 1} (r\|x)^i = (r\|x)^+.
  \end{align*}
\end{proof}

In first-order Kleene algebras (\ref{rg3}) and (\ref{rg4}) are
independent, but it is impossible to find a counterexample with
Nitpick because it generates only finite counterexamples, and all
finite Kleene algebras are a forteriori quantales.

Jones quintuples can be encodeded in this setting as
\begin{align}
r, g \vdash \{p\} c \{q\} \iff p(r\|c) \le q \land c \le g. \label{quin}
\end{align}

\begin{thm}
With this encoding we can derive the standard rely-guarantee inference
rules, as shown in Figure \ref{fig:rgrules}.
\end{thm}

\begin{figure}[hbt]
\centering
\begin{prooftree}
\RightLabel{Skip}
\AxiomC{$pr \le p$}
\UnaryInfC{$r, g \vdash \{p\}1\{p\}$}
\end{prooftree}

@@

Thus equations
(\ref{rg1}) to (\ref{rg4}), which are all necessary to derive these
rules, somehow represent a minimalistic set of axioms from which these
inference rules can be derived.

If we add residuals to our algebra quintuples can be encoded in the
following way, which is equivalent to the encoding in Equation
(\ref{quin}).
\begin{align}
r, g \vdash \{p\} c \{q\} \iff c \le r/(p \rightarrow q) \sqcap g \label{refine}.
\end{align}
This encoding allows us to think in terms of program refinement, as in~\cite{hayes_refining_2013}, since
$r/(p \rightarrow q) \sqcap g$ defines the weakest program that when
placed in parallel with interference from $r$, and guaranteeing
interference at most $g$, goes from $p$ to $q$---a generic
specification for a concurrent program.

@@

While the algebra in section \ref{sec:RG} is adequate for deriving the
standard inference rules, it's equality is too strong to capture many
interesting statements about concurrent programs. Consider the
congruence rule for parallel composition, which is inherent in the
algebraic approach:
\begin{align*}
x = y \implies x\|z = y\|z.
\end{align*}
This can be read as follows; if $x$ and $y$ are equal, then they must
be equal under all possible interference from any arbitrary $z$. At
first sight, this might seem to preclude any reasoning about
interference in a fine-grained way using purely algebra. This is not
the case, but breaking such inherent compositionality in just the
right way to capture interesting properties of interference requires
extra work.

The most obvious way we can acheive this is by expanding our
rely-guarantee algebra with an additional function $\pi$ and redefining
our quintuples as,
\begin{align*}
r, g \vdash \{p\} x \{q\} \iff p(r\|c) \le_\pi q \land x \le g.
\end{align*}
Where $x \le_\pi y$ is $\pi(x) \le \pi(y)$. Since for any operator
$\bullet$, it is not required that
\begin{align*}
\pi(x) = \pi(y) \implies \pi(x \bullet z) = \pi(y \bullet z),
\end{align*}
we can `break' compositionality in just the right way, provided we
choose the correct properties for $\pi$. However, if we simply listed
all properties that $\pi$ needs to satisfy we would end up with an
overly long list of axioms to add to our algebra. Instead we define
the function $\pi$ as $\lambda x.\; x \sqcap c$ where $c$ is the
greatest $consistent$ element such that
\begin{align}
x^* &\le_\pi \pi(x)^*, \label{con1}\\
xy &\le_\pi \pi(x)\pi(y). \label{con2}\\
z + x\cdot y \le_\pi y &\implies x^*\cdot z \le_\pi y\label{con3}\\
z + y\cdot x \le_\pi y &\implies z\cdot x^* \le_\pi y\label{con4}
\end{align}

For any operator $\bullet$, we write $x \bullet_\pi y$ for the
operator $\pi(x\bullet y)$, and we write $x^\pi$ for
$\pi(x^\star)$.
\begin{thm}
Axioms (\ref{con1}) to (\ref{con4}) induce a Kleene subalgebra.
\begin{align*}
(\{x\in K.\; \pi(x) = x\}, +_\pi, \cdot_\pi, \hphantom{}^\pi,0,1)
\end{align*}
\end{thm}
As such we redefine our rely-guarantee algebra from section
\ref{sec:RG} as a structure $(K,RG,+,\sqcap,\cdot,\|,\hphantom{}^\star,0,1,c)$
which, in addition to the rules in section \ref{sec:RG} satisfies both
(\ref{con1}) to (\ref{con4}).
\begin{thm}
All the rules in Figure \ref{fig:rgrules} can be derived with this
algebra, and moreover their proofs remain the same, mutatis
mutandis.
\end{thm}

@@

We have...

\paragraph*{Acknowledgements} This work was supported by funding from
CNPq Brazil and the EPSRC.

\newpage
